Practical Challenges with Web Service Evolution:\\A Review of Literature and Current Industry Practices

### 2013-05-31 paper.tex
references are now using the bibtex format

% a short form should be given in case it is too long for the running head
\titlerunning{Practical Challenges with Web Service Evolution}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{S M Sohan%
\and Frank Maurer}
%
\authorrunning{Practical Challenges with Web Service Evolution}
% (feature abused for this document to repeat the title also on left hand pages)

@@

Web services offer a web-based communication mechanism between two software systems, commonly known as web service publishers and consumers. Wikipedia describes web service as follows:

\begin{quote}
A web service is a method of communication between two electronic devices over the World Wide Web. A web service is a software function provided at a network address over the web or the cloud, it is a service that is ``always on'' as in the concept of utility computing. \[http\:\/\/en.wikipedia.org/wiki/Web_service\]
\end{quote}

@@

Web services provide web-addressable endpoints for the communication between the two parties, where the consumer initiates the requests and the publisher responds to the requests. To carry out the actual communication, the web service publishers and consumers follow specific protocols. Of the many protocols that are in use today \[http://en.wikipedia.org/wiki/List_of_web_service_protocols\], two commonly used protocols are as follows:

@@

\subsection{SOAP Web Services} % (fold)
\label{sub:soap_web_services}
SOAP was designed in 1998 by Dave Winer, Don Box, Bob Atkinson, and Mohsen Al-Ghosein for Microsoft and is currently maintained by the World Wide Web Consortium. [http://en.wikipedia.org/wiki/SOAP] In brief, SOAP is an XML based protocol to communicate structured data and necessary headers between the consumers and publisher of a web service.

@@

The term Representational State Transfer (REST) was introduced by Roy Fielding \[https://en.wikipedia.org/wiki/Representational_state_transfer\#cite_note-Fielding-Ch5-1\] and is used interchangeably as RESTful web service or Web API. In brief, REST is an implementation for web services where user defined ``resources'' are created, accessed, modified and deleted utilizing the various features of the HTTP stack such as caching, security, layering and different request methods (GET, POST, PUT, DELETE etc.), content types, etc.

Due to their historical usage, the term ``web service'' is often interpreted as SOAP. But throughout this paper, unless otherwise specified, we have used this term to represent a web based communication method between two software systems irrespective of their actual implementation details. However, even though there are different kinds of web services \[http://en.wikipedia.org/wiki/Web_service\], the findings mentioned in this paper are based on SOAP and/or REST based services only. This is due to their relative popularity over the other types.
% subsection restful_web_services (end)

@@

\section{Motivation} % (fold)
\label{sec:motivation}
Web services, like other software systems, need to evolve to stay on top of the bugs and ever changing requirements. For example, between May 15th 2013 and May 22nd 2013, the Facebook API team saw 238 new bugs, resolved 211 bugs and fixed 24 bugs. \[https://developers.facebook.com/blog/post/2013/05/22/platform-updates--operation-developer-love/\] However, often times web services are consumed by an unknown number of consumers, and the web service publishers have little control over them. Updating a web service can be tricky in such a situation since a breaking change may upset the existing consumers. To be able to evolve in such a decoupled scenario, multiple versions of a service need to co-exist which poses both technical and business challenges in multiple fronts. At a conceptual level the following diagram shows the different aspects of a web service that need to deal with evolution challenges:

@@

We have performed a Systematic Literature Review to extract information from the published articles following \[Guidelines for performing Systematic Literature Reviews in Software Engineering\]. We have also investigated the published real-word web services to find out the answers to the research questions based on today’s solutions.

\subsection{Systematic Literature Review} % (fold)
\label{sub:systematic_literature_review}
Kitchenham et al defines systematic review as follows [Guidelines for performing Systematic Literature Reviews in Software Engineering]:

@@

To identify the relevant papers from these sources we have performed keyword searches. As mentioned in \[Systematic Mapping Studies in Software Engineering\], keywording is a two step process, 1) extracting keywords from reading the abstracts and, 2) combining keywords from different papers. At step 1, we have initially searched for the papers containing the terms ``web service'' and extracted some commonly used synonyms from their abstracts. We have seen repeated mentions of the terms Service Oriented Architecture or SOA, SOAP, Web API and REST to denote similar concepts. In addition to this, we have expanded the word ``change'' with its commonly used English synonyms. At step 2, we combined the keywords and formed the following case insensitive search query:

@@

We performed searches using this query on the full contents of the articles, ignoring their publication date since the topic of web service is a relatively new one. From the search results, we performed a primary screening after reading the abstracts. Then, for the remaining articles, we continued reading until we could determine if it was relevant or not. If it was found relevant, we read the full contents and extracted information to answer our research questions. For the scope of this research, we defined an article as relevant only if it contained topics about the evolution of web services. After the screening process, we had a shortlist of articles to extract information from. In the next paragraphs, we discuss the findings from these sources in light of the research questions.

@@

\subsubsection{Why Do Web Services Evolve?} % (fold)
\label{sub:why_do_web_services_change}
In this section, we have discussed the drivers behind web service evolution from the related literature. To begin with, in \[On Analyzing Evolutionary Changes of Web Services⋆\] Treiber et al provided a list of changes related to web services with their interdependencies. To identify the source and impact of a change, they categorized the stakeholders into four roles:

@@

Borovskiy et al identified some drivers behind web service evolution from the perspective of an enterprise, where both the web service consumer and publisher may be part of the same organization \[Evolution Management of Enterprise Web Services\]. In this regard, they categorized two classes of drivers behind the evolution of web services as follows:

@@

\subsubsection{Tradeoffs:} % (fold)
Lublinsky compared between multiple approaches of implementing evolving web services based on different dimensions as follows \[Versioning in SOA\].

@@

When multiple versions of a service are deployed at the same time, a versioning scheme is required to identify and use a specific version. Leskey discusses about the versioning of web services in \[Considerations for Versioning SOA Resources\] with respect to the OASIS SOA Reference Model (SOA-RM). In this paper, the author discussed versions in terms of identifiable resources, that can have multiple versions and are modified by applying new versions to the original one.

@@

\subsubsection{Protocol Level Implementation:}
Since web services are typically based on protocols such as SOAP, REST, etc., protocol level support for evolving web services would allow for uniform standards. In \[Semantically Extensible Schemas for Web Service Evolution\] Wilde proposed a conceptual framework for SOAP web services where consumers can be forward compatible in a managed way.

@@

In \[End-to-End Versioning Support for Web Services\] Leitner et al provided an alternative approach to transparently handle multiple versions of SOAP web services using their framework called VRESCo. At it’s core, VRESCo builds a graph of the versions showing predecessor relations as follows:

@@

In addition to the protocol level evolution support for SOAP web services, Mangler et al showed a solution for RESTful web services in \[On the Origin of Services - Using RIDDL for Description, Evolution and Composition of RESTful Services\]. They introduced a description language called RIDDL for RESTful web services to support service composition and evolution.

@@

Kaminski et al presented a design technique for evolving web services from the perspective of the source code implementation \[A Design Technique for Evolving Web Services\]. They identified some key requirements for their design such as: a) backward compatibility, b) common data store, c) no code duplication, d) untangled versions, e) unconstrained evolution, and f) a visible mechanism. To summarize, the design aimed to allow for web service evolution while staying compatible with the existing clients without introducing duplication in the code or in the data store in a transparent manner.

@@

\[Source: A Design Technique for Evolving Web Services \]

@@

In \[On Synchronizing with Web Service Evolution\] Zou et al presents a technique to generate consumer specific changelogs for evolving SOAP web services. When a new version of a web service is published, it typically accompanies a custom formatted, often HTML based, documentation of it’s methods and parameters. For existing consumers, this may require a significant effort to identify the impacts of the new version specific to their usage by looking at the potentially large documentation. To address this problem, the authors suggested creating customized release notes for each consumer based on their usage pattern in the following steps:

@@

Aversano et al provided an alternate approach to communicate changes in SOAP web services using a graphical way in \[Visualizing the Evolution of Web Services using Formal Concept Analysis\]. To construct the graphical representation, the authors extracted meta information about methods and parameters for SOAP web services from the WSDL files. This meta information is then presented in a graph to visualize the service with its methods and parameters. The graph can be used to visually communicate the similarities and differences among the subsequent versions of the same service.

@@

\subsubsection{Google Maps:}
Google Maps provides a JavaScript based web service for it’s consumers and updates the service to provide ``new features, bug fixes and performance improvements'' \[https://developers.google.com/maps/documentation/javascript/basics\#Versioning\]. Their versioning policy offers the following versions:

@@

\subsubsection{Twitter API:}
Twitter provides a RESTful web service to interact with it’s data from third-party applications \[https://dev.twitter.com/docs/api\]. This web service is also evolving to fix bugs and bring new features, but using a different approach from that of Google Maps.

At the time of writing this paper, they have versions 1 and 1.1 deployed for the consumers. Version 1 is marked as deprecated and announced to be retired in six months after version 1.1 was released\[https://dev.twitter.com/docs/api/1.1/overview\]. The consumers can specify the target version in the URL. For example, the web service calls to https://api.twitter.com/1.1/ will use version 1.1.

Once a version is released, future changes are made to the released version to provide new features and bug fixes. However, unlike Google Maps, new features are not always released under a new version name. Instead, the changes may be deployed under the same version name and are communicated to the consumers using an updated “Calendar of API Changes” \[https://dev.twitter.com/calendar\] through a developer portal.

The developer portal is also used to record issues and collect feedback from the consumers. The portal contains version specific HTML based documentations of the web service as well as an interactive web browser based console to explore the service on the browser \[https://dev.twitter.com/console\].

@@

Unlike the previous approaches, Facebook Graph API on the other hand uses a different approach for it’s evolution. Facebook Graph API is the primary web service to retrieve and post data to Facebook \[https://developers.facebook.com/docs/reference/api/\]. They have a 90-day breaking change policy stated as follows:

@@

To communicate the web service changes with it’s consumers and collect feedback, Facebook uses a developer portal as well as a StackOverflow channel \[http://facebook.stackoverflow.com/\]. The developer portal shares information about planned upcoming changes as well as the release notes of each version.

@@

\begin{thebibliography}{4}

\bibitem{jour} Smith, T.F., Waterman, M.S.: Identification of Common Molecular
Subsequences. J. Mol. Biol. 147, 195--197 (1981)

\bibitem{lncschap} May, P., Ehrlich, H.C., Steinke, T.: ZIB Structure Prediction Pipeline:
Composing a Complex Biological Workflow through Web Services. In: Nagel,
W.E., Walter, W.V., Lehner, W. (eds.) Euro-Par 2006. LNCS, vol. 4128,
pp. 1148--1158. Springer, Heidelberg (2006)

\bibitem{book} Foster, I., Kesselman, C.: The Grid: Blueprint for a New Computing
Infrastructure. Morgan Kaufmann, San Francisco (1999)

\bibitem{proceeding1} Czajkowski, K., Fitzgerald, S., Foster, I., Kesselman, C.: Grid
Information Services for Distributed Resource Sharing. In: 10th IEEE
International Symposium on High Performance Distributed Computing, pp.
181--184. IEEE Press, New York (2001)

\bibitem{proceeding2} Foster, I., Kesselman, C., Nick, J., Tuecke, S.: The Physiology of the
Grid: an Open Grid Services Architecture for Distributed Systems
Integration. Technical report, Global Grid Forum (2002)

\bibitem{url} National Center for Biotechnology Information, \url{http://www.ncbi.nlm.nih.gov}

\end{thebibliography}

>>>>

% a short form should be given in case it is too long for the running head
\titlerunning{Challenges with Web Service Evolution}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{S M Sohan%
\and Frank Maurer}
%
\authorrunning{Challenges with Web Service Evolution}
% (feature abused for this document to repeat the title also on left hand pages)

@@

Web services offer a web-based communication mechanism between two software systems, commonly known as web service publishers and consumers. Wikipedia describes web service as follows \cite{web_service_wiki}:

\begin{quote}
A web service is a method of communication between two electronic devices over the World Wide Web. A web service is a software function provided at a network address over the web or the cloud, it is a service that is ``always on'' as in the concept of utility computing.
\end{quote}

@@

Web services provide web-addressable endpoints for the communication between the two parties, where the consumer initiates the requests and the publisher responds to the requests. To carry out the actual communication, the web service publishers and consumers follow specific protocols. Of the many protocols that are in use today \cite{web_service_protocols_wiki}, two commonly used protocols are as follows:

@@

\subsection{SOAP Web Services} % (fold)
\label{sub:soap_web_services}
SOAP was designed in 1998 by Dave Winer, Don Box, Bob Atkinson, and Mohsen Al-Ghosein for Microsoft and is currently maintained by the World Wide Web Consortium \cite{soap_wiki}. In brief, SOAP is an XML based protocol to communicate structured data and necessary headers between the consumers and publisher of a web service.

@@

The term Representational State Transfer (REST) was introduced by Roy Fielding \cite{rest_wiki} and is used interchangeably as RESTful web service or Web API. In brief, REST is an implementation for web services where user defined ``resources'' are created, accessed, modified and deleted utilizing the various features of the HTTP stack such as caching, security, layering and different request methods (GET, POST, PUT, DELETE etc.), content types, etc.

Due to their historical usage, the term ``web service'' is often interpreted as SOAP. But throughout this paper, unless otherwise specified, we have used this term to represent a web based communication method between two software systems irrespective of their actual implementation details. However, even though there are different kinds of web services, the findings mentioned in this paper are based on SOAP and/or REST based services only. This is due to their relative popularity over the other types.
% subsection restful_web_services (end)

@@

\section{Motivation} % (fold)
\label{sec:motivation}
Web services, like other software systems, need to evolve to stay on top of the bugs and ever changing requirements. For example, between May 15th 2013 and May 22nd 2013, the Facebook API team saw 238 new bugs, resolved 211 bugs and fixed 24 bugs \cite{facebook_release_note}. However, often times web services are consumed by an unknown number of consumers, and the web service publishers have little control over them. Updating a web service can be tricky in such a situation since a breaking change may upset the existing consumers. To be able to evolve in such a decoupled scenario, multiple versions of a service need to co-exist which poses both technical and business challenges in multiple fronts. At a conceptual level the following diagram shows the different aspects of a web service that need to deal with evolution challenges:

@@

We have performed a Systematic Literature Review to extract information from the published articles following \cite{kitchenham2007guidelines}. We have also investigated the published real-word web services to find out the answers to the research questions based on today’s solutions.

\subsection{Systematic Literature Review} % (fold)
\label{sub:systematic_literature_review}
Kitchenham et al defines systematic review as follows:

@@

To identify the relevant papers from these sources we have performed keyword searches. As mentioned in \cite{petersen2008systematic}, keywording is a two step process, 1) extracting keywords from reading the abstracts and, 2) combining keywords from different papers. At step 1, we have initially searched for the papers containing the terms ``web service'' and extracted some commonly used synonyms from their abstracts. We have seen repeated mentions of the terms Service Oriented Architecture or SOA, SOAP, Web API and REST to denote similar concepts. In addition to this, we have expanded the word ``change'' with its commonly used English synonyms. At step 2, we combined the keywords and formed the following case insensitive search query:

@@

We performed searches using this query on the full contents of the articles, ignoring their publication date since the topic of web service is a relatively new one. From the search results, we performed a primary screening after reading the abstracts. Then, for the remaining articles, we continued reading until we could determine if it was relevant or not. If it was found relevant, we read the full contents and extracted information to answer our research questions. For the scope of this research, we defined an article as relevant only if it contained topics about the evolution of web services.

After the screening process, we had a shortlist of articles to extract information from. In the next paragraphs, we discuss the findings from these sources in light of the research questions.

@@

\subsubsection{Why Do Web Services Evolve?} % (fold)
\label{sub:why_do_web_services_change}
In this section, we have discussed the drivers behind web service evolution from the related literature. To begin with, in \cite{treiber2009analyzing} Treiber et al provided a list of changes related to web services with their interdependencies. To identify the source and impact of a change, they categorized the stakeholders into four roles:

@@

Borovskiy et al identified some drivers behind web service evolution from the perspective of an enterprise, where both the web service consumer and publisher may be part of the same organization \cite{borovskiy2008evolution}. In this regard, they categorized two classes of drivers behind the evolution of web services as follows:

@@

\subsubsection{Tradeoffs:} % (fold)
Lublinsky compared between multiple approaches of implementing evolving web services based on different dimensions as follows \cite{lublinsky2007versioning}.

@@

When multiple versions of a service are deployed at the same time, a versioning scheme is required to identify and use a specific version. Leskey discusses about the versioning of web services in \cite{laskey2008considerations} with respect to the OASIS SOA Reference Model (SOA-RM). In this paper, the author discussed versions in terms of identifiable resources, that can have multiple versions and are modified by applying new versions to the original one.

@@

\subsubsection{Protocol Level Implementation:}
Since web services are typically based on protocols such as SOAP, REST, etc., protocol level support for evolving web services would allow for uniform standards. In \cite{wilde2004semantically} Wilde proposed a conceptual framework for SOAP web services where consumers can be forward compatible in a managed way.

@@

In \cite{leitner2008end} Leitner et al provided an alternative approach to transparently handle multiple versions of SOAP web services using their framework called VRESCo. At it’s core, VRESCo builds a graph of the versions showing predecessor relations as follows:

@@

In addition to the protocol level evolution support for SOAP web services, Mangler et al showed a solution for RESTful web services in \cite{mangler2010origin}. They introduced a description language called RIDDL for RESTful web services to support service composition and evolution.

@@

Kaminski et al presented a design technique for evolving web services from the perspective of the source code implementation \cite{kaminski2006design}. They identified some key requirements for their design such as: a) backward compatibility, b) common data store, c) no code duplication, d) untangled versions, e) unconstrained evolution, and f) a visible mechanism. To summarize, the design aimed to allow for web service evolution while staying compatible with the existing clients without introducing duplication in the code or in the data store in a transparent manner.

@@

Source: \cite{kaminski2006design}

@@

In \cite{le2008synchronizing} Zou et al presents a technique to generate consumer specific changelogs for evolving SOAP web services. When a new version of a web service is published, it typically accompanies a custom formatted, often HTML based, documentation of it’s methods and parameters. For existing consumers, this may require a significant effort to identify the impacts of the new version specific to their usage by looking at the potentially large documentation. To address this problem, the authors suggested creating customized release notes for each consumer based on their usage pattern in the following steps:

@@

Aversano et al provided an alternate approach to communicate changes in SOAP web services using a graphical way in \cite{aversano2005visualizing}. To construct the graphical representation, the authors extracted meta information about methods and parameters for SOAP web services from the WSDL files. This meta information is then presented in a graph to visualize the service with its methods and parameters. The graph can be used to visually communicate the similarities and differences among the subsequent versions of the same service.

@@

\subsubsection{Google Maps:}
Google Maps provides a JavaScript based web service for it’s consumers and updates the service to provide ``new features, bug fixes and performance improvements'' \cite{google_maps_versioning}. Their versioning policy offers the following versions:

@@

\subsubsection{Twitter API:}
Twitter provides a RESTful web service to interact with it’s data from third-party applications \cite{twitter_api}. This web service is also evolving to fix bugs and bring new features, but using a different approach from that of Google Maps.

At the time of writing this paper, they have versions 1 and 1.1 deployed for the consumers. Version 1 is marked as deprecated and announced to be retired in six months after version 1.1 was released. The consumers can specify the target version in the URL. For example, the web service calls to https://api.twitter.com/1.1/ will use version 1.1.

Once a version is released, future changes are made to the released version to provide new features and bug fixes. However, unlike Google Maps, new features are not always released under a new version name. Instead, the changes may be deployed under the same version name and are communicated to the consumers using an updated “Calendar of API Changes” \cite{twitter_calendar} through a developer portal.

The developer portal is also used to record issues and collect feedback from the consumers. The portal contains version specific HTML based documentations of the web service as well as an interactive web browser based console to explore the service on the browser \cite{twitter_console} .

@@

Unlike the previous approaches, Facebook Graph API on the other hand uses a different approach for it’s evolution. Facebook Graph API is the primary web service to retrieve and post data to Facebook \cite{facebook_api}. They have a 90-day breaking change policy stated as follows:

@@

To communicate the web service changes with it’s consumers and collect feedback, Facebook uses a developer portal as well as a StackOverflow channel \cite{facebook_stack_overflow}. The developer portal shares information about planned upcoming changes as well as the release notes of each version.

@@

\bibliographystyle{unsrt}
\bibliography{references}
